---
title: "ARTEMIS Installation Guide"
author: "Strahinja Stevanovic, Svetozar Nesic"
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{ARTEMIS Installation Guide}
  %\VignetteEncoding{UTF-8}
---
# Introduction

This vignette describes how you need to install the Observational Health Data Science and Informatics (OHDSI) [`ARTEMIS`](https://github.com/OHDSI/ARTEMIS) package under Windows, Mac, and Linux.

# Software Prerequisites

Before installing ARTEMIS, ensure that **Python (version ≥ 3.12)** is installed on your system.  
You can check which Python version R detects using:

```{r, echo = TRUE, message = FALSE, warning = FALSE,tidy=FALSE,eval=FALSE}
system("python --version", intern = TRUE)
```

If you want ARTEMIS to use a specific Python interpreter, set the `ARTEMIS_PYTHON` environment variable before installation:

```{r, echo = TRUE, message = FALSE, warning = FALSE,tidy=FALSE,eval=FALSE}
Sys.setenv(ARTEMIS_PYTHON = "/path/to/your/python")
```

ARTEMIS can be installed directly from GitHub:

```{r, echo = TRUE, message = FALSE, warning = FALSE,tidy=FALSE,eval=FALSE}
# Install devtools if it is not already installed
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

# Install ARTEMIS from GitHub
devtools::install_github("OHDSI/ARTEMIS")
```

If you are unsure how to install `Python` or set `ARTEMIS_PYTHON`, refer to the OS-specific setup instructions below.

### On Windows
Install Python 3.12 or above from the Microsoft Store or from:
https://www.python.org/downloads/windows/

Then open cmd or PowerShell and set the environment variable:

    set ARTEMIS_PYTHON=<ABS\\PATH\\TO\\python.exe (v3.12+)>

Other requirements checklist:

* R, Rtools, and devtools installed

* Microsoft Visual C++ 14.0 or greater (required by Python packages like numpy)

* Visual Studio Build Tools (for faster Cython-compiled alignment)

### On Linux or macOS

Install Python 3.12+ using your preferred package manager (e.g., Homebrew, apt, pacman) or download it from: https://www.python.org/

Then from the terminal, set the Python version environment variable:

    export ARTEMIS_PYTHON="/absolute/path/to/python3.12"

Other dependencies you might need installed:

    base-devel, r, git, libgit2, zlib, libxml2, openssl, curl, pkgconf, 
    pandoc, glpk, gmp, libtool, graphviz, make, cmake, tzdata, 
    jdk-openjdk, libcurl-compat, gcc-fortran, openblas, lapack

### Reticulate

💡 You do NOT need to manually set up reticulate — ARTEMIS takes care of it automatically during setup. This section is for informational purposes only.

ARTEMIS relies on a python back-end via
[reticulate](https://rstudio.github.io/reticulate/) and depending on
your reticulate settings, system and environment, you may need to run
the following commands before loading the package:

    reticulate::py_install("numpy")
    reticulate::py_install("pandas")

    **Other python dependencies for the build**
    reticulate::py_install("setuptools")
    reticulate::py_install("wheel")
    reticulate::py_install("Cython")
    reticulate::py_install("tqdm")
    
If you do not presently have reticulate or python3.12 installed you may
first need to run the following commands to ensure that reticulate can
access a valid python install on your system:

    install.packages("reticulate")
    library(reticulate)

This will prompt reticulate to install python, create a local virtualenv
called “r-reticulate” and, finally, set this virtual environment as the
local environment for use when running python via R through reticulate.
